Do investors in clean energy ETFs herd? The role of climate risks

### 2025-06-26 Journal/journal_submission/alt_energy_herding_full.tex
updated version to journal

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrartcl}

\usepackage{amsmath,amssymb}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\usepackage{fontspec}
\usepackage{multirow}
\usepackage{multicol}
\usepackage{colortbl}
\usepackage{hhline}
\newlength\Oldarrayrulewidth
\newlength\Oldtabcolsep
\usepackage{longtable}
\usepackage{array}
\usepackage{hyperref}
\usepackage{float}
\usepackage{wrapfig}
% \usepackage{hyperref}
% \hypersetup{
%     colorlinks,
%     linkcolor={blue!100!black},
%     citecolor={blue!100!black},
%     urlcolor={blue!100!black}
% }
% \usepackage{apacite}
% \usepackage[round]{natbib} 
% \usepackage{graphicx}
% \usepackage{float}
% \usepackage{caption}
% \usepackage[toc,page]{appendix}
% \usepackage{booktabs,caption}
% \usepackage[flushleft]{threeparttable}
% \usepackage{tabularx}
\usepackage{sfmath}
\usepackage{fontspec}
\usepackage[utf8]{inputenc}
\usepackage{siunitx}
\usepackage{amsfonts}
\usepackage{amsmath}
% \usepackage{xcolor}
% \usepackage{scrextend}
% \deffootnote[2em]{2em}{1em}{\textsuperscript{\thefootnotemark}\,}
% \newcolumntype{Y}{>{\centering\arraybackslash}X}
% \usepackage[a4paper]{geometry}
% \usepackage{caption}
% \usepackage[bottom,flushmargin,hang,multiple]{footmisc}
% \usepackage{pdflscape}
% \usepackage[paper=portrait,pagesize]{typearea}
% \usepackage{rotating}
% \usepackage{fullpage}
% \usepackage{tabu}
\usepackage{lscape}
\newcommand{\blandscape}{\begin{landscape}}
\newcommand{\elandscape}{\end{landscape}}
\usepackage{rotating}
\newcommand{\bsideways}{\begin{sidewaystable}[htbp]}
\newcommand{\esideways}{\end{sidewaystable}}
\KOMAoption{captions}{tableheading}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[round]{natbib}
\bibliographystyle{apalike}
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  colorlinks=true,
  linkcolor={blue!100!black},
  filecolor={Maroon},
  citecolor={blue!100!black},
  urlcolor={blue!100!black},
  pdfcreator={LaTeX via pandoc}}

\author{}
\date{}

\begin{document}

@@

Exchange-traded funds (ETFs) are a key feature of green financial
products. They are a collection of investments that often tracks an
underlying performance of an asset or index. In the universe of
environment, social, and governance (ESG) investments, clean ETFs serve
as a tool for environmentally conscious investors to identify and invest
in environmentally friendly companies \citep{briere2023} engaged in the
transition towards a cleaner production and a low-carbon economy. Among
the ESG ETFs, the Clean Energy (CE) ETFs have been the best-performing
one in 2022 \citep{decclesia2024}. This is not surprising given that the
clean energy transition represents one of the largest multi-decade
secular growth opportunities. After the inclusion of Green energy
financing in the list of United Nations Sustainability Goals (SDGs) as
SDG 7, the role, importance, and visibility of green financial products
have escalated enormously \citep{naqvi2022}. That is, the growth of
green assets under management is likely to continue. Furthermore, the
limited availability of data on ESG complying investment tools
\citep{avramov2022, nguyen2025} justifies the use of energy ETFs as best
candidates of green assets. However, given that the inclusion of climate
sustainability in investment decisions is a recent phenomenon, it is not
clear what the actual impact will be in the long-run.

@@

Following the relevant literature
\citetext{\citealp{christie1995}; \citealp[and][]{chang2000}}, we
compute dispersion of the \(i^{th}\) ETF from the market return, which
is known as the Cross Sectional Absolute Deviation (\(CSAD_t\)) measure.
Empirically the \(CSAD_t\) is defined in the following manner:

@@

>>>>

% Options for packages loaded elsewhere
% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrartcl}
\usepackage{xcolor}
\usepackage{amsmath,amssymb}
\setcounter{secnumdepth}{5}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{setspace}
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother

\setlength{\emergencystretch}{3em} % prevent overfull lines

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage[round]{natbib}
\bibliographystyle{apalike}

\usepackage{fontspec}
\usepackage{multirow}
\usepackage{multicol}
\usepackage{colortbl}
\usepackage{hhline}
\newlength\Oldarrayrulewidth
\newlength\Oldtabcolsep
\usepackage{longtable}
\usepackage{array}
\usepackage{hyperref}
\usepackage{float}
\usepackage{wrapfig}
% \usepackage{hyperref}
% \hypersetup{
%     colorlinks,
%     linkcolor={blue!100!black},
%     citecolor={blue!100!black},
%     urlcolor={blue!100!black}
% }
% \usepackage{apacite}
% \usepackage[round]{natbib} 
% \usepackage{graphicx}
% \usepackage{float}
% \usepackage{caption}
% \usepackage[toc,page]{appendix}
% \usepackage{booktabs,caption}
% \usepackage[flushleft]{threeparttable}
% \usepackage{tabularx}
\usepackage{sfmath}
\usepackage{fontspec}
\usepackage[utf8]{inputenc}
\usepackage{siunitx}
\usepackage{amsfonts}
\usepackage{amsmath}
% \usepackage{xcolor}
% \usepackage{scrextend}
% \deffootnote[2em]{2em}{1em}{\textsuperscript{\thefootnotemark}\,}
% \newcolumntype{Y}{>{\centering\arraybackslash}X}
% \usepackage[a4paper]{geometry}
% \usepackage{caption}
% \usepackage[bottom,flushmargin,hang,multiple]{footmisc}
% \usepackage{pdflscape}
% \usepackage[paper=portrait,pagesize]{typearea}
% \usepackage{rotating}
% \usepackage{fullpage}
% \usepackage{tabu}
\usepackage{lscape}
\newcommand{\blandscape}{\begin{landscape}}
\newcommand{\elandscape}{\end{landscape}}
\usepackage{rotating}
\newcommand{\bsideways}{\begin{sidewaystable}[htbp]}
\newcommand{\esideways}{\end{sidewaystable}}
\KOMAoption{captions}{tableheading}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  colorlinks=true,
  linkcolor={blue!100!black},
  filecolor={Maroon},
  citecolor={blue!100!black},
  urlcolor={blue!100!black},
  pdfcreator={LaTeX via pandoc}}

\author{}
\date{}
\begin{document}

@@

Energy market has been in the epicentre of retail investors and
portfolio managers at least over the last two decades. Official
estimates claim that global energy investment is expected to surpass USD
3 trillion for the first time in 2024, with USD 2 trillion going to
clean energy technologies and infrastructure \citep{iea2024}. There is a
general consensus in the literature that the most important event that
has shaped the current landscape in the energy sector is the emergence
of the energy portfolio with the rise of green energy as an alternative
energy resource and an asset class. Price volatility of fossil fuels on
the one hand and decarbonisation-related goals dictated by international
agreements on the other hand are heavily responsible for the growing
interest in green energy market. Responding to the increasing awareness
about the environment and the need for more green capital, financial
markets have also attracted a large portion of retail and institutional
investors' fund flows \citep{rizvi2022green, naqvi2021there}.

Needless to mention that the rise of energy popularity as an alternative
investment class has also been fueled by the fictionalization of
commodities markets. Commodities have become an effective investment
tool for investment community as it allows investors to participate in
the market developments across various physical products such as
agricultural products, silver, gold, natural gas, oil etc and more
recently clean energy products without being obliged to own the product
directly \citep{gagnon2020they}. Several academic studies have
highlighted substantial diversification benefits of commodities when
added to a portfolio of stocks and bonds \citep[see inter
alia,][]{naqvi2022}.

Green energy sectors like solar, wind, geothermal, biofuels, biomass,
wave, hydro and tidal energies are considered by investors, authorities
and policy-makers as the substitutes for fossil-fuels energy that
increase \(CO_2\) emissions \citep{shahzad2020clean}. Moreover, as
\citet{naqvi2022} point out that ``the alternative energy sector is
expected to grow faster over the next decades due to the depletion of
fossil fuel reserves and the technological advance that has made green
energy feasible''.

Although green energy investment has witnessed an exponential growth
over the years, there is a lot of improvement that can be done in terms
of investors' participation, both retail and institutional. To achieve
the goal of larger capital commitment to green energy projects it is
important to gain investors' confidence in financial products that are
linked to green energy such as Exchange-traded funds(ETFs).

ETFs are a key feature of green financial products. They are a
collection of investments that often tracks an underlying performance of
an asset or index. In the universe of environment, social, and
governance (ESG) investments, clean ETFs serve as a tool for
environmentally conscious investors to identify and invest in
environmentally friendly companies \citep{briere2023} engaged in the
transition towards a cleaner production and a low-carbon economy. Among
the ESG ETFs, the Clean Energy (CE) ETFs have been the best-performing
one in 2022 \citep{decclesia2024}. This is not surprising given that the
clean energy transition represents one of the largest multi-decade
secular growth opportunities. After the inclusion of Green energy
financing in the list of United Nations Sustainability Goals (SDGs) as
SDG 7, the role, importance, and visibility of green financial products
have escalated enormously \citep{naqvi2022}. That is, the growth of
green assets under management is likely to continue. Furthermore, the
limited availability of data on ESG complying investment tools
\citep{avramov2022, nguyen2025} justifies the use of energy ETFs as best
candidates of green assets. However, given that the inclusion of climate
sustainability in investment decisions is a recent phenomenon, it is not
clear what the actual impact will be in the long-run.

@@

More broadly, studies that focus on herding can be broadly classified
into two main categories: on the one hand there are studies that examine
institutional investor herding and analyst herding using
transaction-level data
\citetext{\citealp{lakonishok1992}; \citealp[and][among
others]{sias2004institutional}}. On the other hand, there are studies
that explore herding towards the market consensus employing individual
returns and market returns
\citetext{\citealp{christie1995}; \citealp[and][]{chang2000}}. In the
latter case, the most widely employed method is the cross-sectional
standard deviation of returns proposed by \citet{christie1995}. The
intuition behind this approach is that herding toward the market is
signaled when return dispersion is relatively low. Rational asset
pricing theory predicts that when stocks respond to market movements in
various degrees then dispersion will be linearly related to market
returns \citep[see also,][]{hwang2004market}. However, when investors
tend to mimic each other and herd towards the market the linear relation
between returns dispersion and aggregate market returns is no longer
valid. Under these conditions, fitting a non-linear regression
specification between a measure of returns dispersion and aggregate
market return appears as a more viable choice. Therefore, our approach
to measure herding activity is based on the non-linear regression model
by \citet{chang2000}.

Following this relevant literature, we compute dispersion of the
\(i^{th}\) ETF from the market return, which is known as the Cross
Sectional Absolute Deviation (\(CSAD_t\)) measure. Empirically the
\(CSAD_t\) is defined in the following manner:

@@

Furthermore, higher transition risk means greater uncertainty for future
pathway towards a cleaner economy. Under these conditions, green assets
such as green stocks or ETFs are expected to perform better and can
serve as hedging tools \citep[see inter alia][]{pastor}. As a result,
investors allocate their capital to available green assets thus
resulting in higher returns' dispersion with respect to the aggregate
market. Stated differently, investors move their capital away from
market consensus across green ETFs ruling out the possibility of herding
behavior.

### 2025-06-26 Journal/journal_submission/alt_energy_herding_no_authors.tex
updated version to journal

\author{ 
Vassilios Babalos\footnote{Corresponding Author. Department of Accounting and Finance, University of Peloponnese, Kalamata,  Greece; Email: v.babalos@uop.gr.} \,\,
Xolani Sibande\footnote{Department of Economics, University of Pretoria, Pretoria, South Africa; Email: xolaniss@gmail.com. Economic Research Department, South African Reserve Bank, Pretoria, South Africa.} \,\, 
Elie Bouri\footnote{Adnan Kassar School of Business, Lebanese American University, Lebanon; Email: elie.elbouri@lau.edu.lb} \,\,
Rangan Gupta\footnote{Department of Economics, University of Pretoria, Pretoria, South Africa; Email: rangan.gupta@up.ac.za.} 
}
\date{\today}
\maketitle

>>>>

%\author{ 
%Vassilios Babalos\footnote{Corresponding Author. Department of Accounting and Finance, University of Peloponnese, Kalamata,  Greece; Email: v.babalos@uop.gr.} \,\,
%Xolani Sibande\footnote{Department of Economics, University of Pretoria, Pretoria, South Africa; Email: xolaniss@gmail.com. Economic Research Department, South African Reserve Bank, Pretoria, South Africa.} \,\, 
%Elie Bouri\footnote{Adnan Kassar School of Business, Lebanese American University, Lebanon; Email: elie.elbouri@lau.edu.lb} \,\,
%Rangan Gupta\footnote{Department of Economics, University of Pretoria, Pretoria, South Africa; Email: rangan.gupta@up.ac.za.} 
%}
\date{\today}
\maketitle

### 2025-06-26 alt_energy_herding.qmd
updated version to journal

Green energy sectors like solar, wind, geothermal, biofuels, biomass, wave, hydro and tidal energies are considered by investors, authorities and policy-makers as the substitutes for fossil-fuels energy that increase CO2 emissions [@shahzad2020clean]. Moreover, as @naqvi2022 point out that "the alternative energy sector is expected to grow faster over the next decades due to the depletion of fossil fuel reserves and the technological advance that has made green energy feasible".

>>>>

Green energy sectors like solar, wind, geothermal, biofuels, biomass, wave, hydro and tidal energies are considered by investors, authorities and policy-makers as the substitutes for fossil-fuels energy that increase $CO_2$ emissions [@shahzad2020clean]. Moreover, as @naqvi2022 point out that "the alternative energy sector is expected to grow faster over the next decades due to the depletion of fossil fuel reserves and the technological advance that has made green energy feasible".
